---
title: "Antimicrobial treatment clearance analysis"
author: "Gerry Tonkin-Hill"
date: "`r Sys.Date()`"
output: 
  html_document:
    fig_width: 12
    fig_height: 8
editor_options: 
  chunk_output_type: console
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(fig.width=20, fig.height=12,
                      echo=TRUE, warning=FALSE, message=FALSE,
                      tidy=TRUE)
options(stringsAsFactors = FALSE)
```
  
##Load Libraries

```{r}
library(data.table)
library(tidyverse)
library(broom)
library(ggthemes)
library(patchwork)
```

##Load epidemiological data and CDC resistance calls

```{r}
CDC_resistance <- fread("./data/TABLE_SPN_CDC_v1.0.3_output_20200527_Typing_Results.txt", 
                        data.table = FALSE, sep="\t", fill=TRUE) %>% as_tibble()
manifests_qc <- fread("./data/manifest_w_QC.csv", data.table = FALSE) %>% as_tibble()
ena_accessesions <- fread("./data/pneumoDEEP_accessions.csv", data.table = FALSE, 
                          header = FALSE, col.names = c("lane", "ena_acc")) %>% as_tibble()
swab_summary <- fread("./data/epi/Pneumo-DEEP_swab_summary.csv", data.table = FALSE) %>% as_tibble
abx_summary <- fread("./data/epi/Pneumo-DEEP_illness_abx_summary.csv", data.table = FALSE) %>% as_tibble
merged_serotype_calls <- fread("./data/merged_serotype_calls.csv", data.table = FALSE) %>% as_tibble()
library_stats <- fread("./data/5000.pathfind_stats.csv", data.table = FALSE) %>% as_tibble()

conservative_carriage_estimates <- fread("./data/carriage_episode_estimates.csv", data.table = FALSE) %>% as_tibble()

merged_serotype_calls$carriage_episode <- conservative_carriage_estimates$carriage_episode[
  match(paste(merged_serotype_calls$specnum, merged_serotype_calls$GPSC),
        paste(conservative_carriage_estimates$specnum, conservative_carriage_estimates$GPSC))
]

CDC_resistance$lane <- gsub("_[0-9]*$", "", CDC_resistance$Sample)
CDC_resistance$GPSC <- as.numeric(gsub(".*_", "", CDC_resistance$Sample))
temp <- ena_accessesions$lane[match(CDC_resistance$lane, ena_accessesions$ena_acc)]
CDC_resistance$lane[!is.na(temp)] <- temp[!is.na(temp)]
CDC_resistance$specnum <- manifests_qc$`DONOR ID (required for EGA)`[match(CDC_resistance$lane, manifests_qc$`Lane Name`)]

CDC_resistance$consensus_percentage <- merged_serotype_calls$consensus_percentage[
  match(paste(CDC_resistance$specnum,
              CDC_resistance$GPSC),
        paste(merged_serotype_calls$specnum,
              merged_serotype_calls$GPSC))]

CDC_resistance$treated <- merged_serotype_calls$treatment[
  match(paste(CDC_resistance$specnum, CDC_resistance$GPSC),
        paste(merged_serotype_calls$specnum,
              merged_serotype_calls$GPSC))]

CDC_resistance$host <- merged_serotype_calls$host[
  match(paste(CDC_resistance$specnum, CDC_resistance$GPSC),
        paste(merged_serotype_calls$specnum,
              merged_serotype_calls$GPSC))]

CDC_resistance <- CDC_resistance %>% 
  filter(paste(specnum, GPSC) %in% 
           paste(merged_serotype_calls$specnum, merged_serotype_calls$GPSC))
```

Initially we assign each swab as being pre or post treatment if the swab was taken within 4 weeks of an antibiotic treatment event. Similarly we associate the diagnosis of within 4 weeks of a swab.

```{r}
swab_summary$specdate <- lubridate::as_date(swab_summary$specdate)
swab_summary$ID <- paste(swab_summary$codenum, swab_summary$category, sep = "-")

#map samples with treatment status
abx_summary$ID <- paste(abx_summary$codenum, "Infant", sep="-")
abx_summary$date_start <- lubridate::as_date(abx_summary$date_start)

#add treatment information
swab_summary$treatment <- map_chr(1:nrow(swab_summary), function(i){
  sampleID <- swab_summary$ID[[i]]
  treatments <- abx_summary[abx_summary$ID==sampleID,]
  treatments <- treatments[!is.na(treatments$drug1),]
  diff <- swab_summary$specdate[[i]]-treatments$date_start
  if (any((diff>0) & (diff<=28))){
    return("TREATED")
  } else if (any((diff>0) & (diff<60))) {
    return("AMBIGUOUS")
  } else {
    return("NOT_TREATED")
  }
})

#add diagnosis information
swab_summary$diagnosis <- map(1:nrow(swab_summary), function(i){
  sampleID <- swab_summary$ID[[i]]
  treatments <- abx_summary[abx_summary$ID==sampleID,]
  treatments <- treatments[!is.na(treatments$drug1),]
  treatments <- treatments[abs(swab_summary$specdate[[i]]-treatments$date_start)<28,, drop=FALSE]
  if (nrow(treatments)>0){
    return(unique(treatments$diagnosis_clinical))
  } else {
    return(NA)
  }
})

swab_summary <- swab_summary[!duplicated(swab_summary$specnum),]
merged_serotype_calls$treatment <- swab_summary$treatment[match(merged_serotype_calls$specnum, swab_summary$specnum)]
merged_serotype_calls$diagnosis <- swab_summary$diagnosis[match(merged_serotype_calls$specnum, swab_summary$specnum)]
merged_serotype_calls$host <- swab_summary$ID[match(merged_serotype_calls$specnum, swab_summary$specnum)]
```

We now identify pairs of sample before and after antimicrobial treatment and corresponding samples where no treatment has occurred.

```{r}
#First identify pairs of samples from the same patient that are within 6 weeks of each other.
test_pairs <- map_dfr(unique(swab_summary$ID), function(id){
  idswabs <- swab_summary[swab_summary$ID==id, , drop=FALSE]
  idswabs <- idswabs[!duplicated(idswabs$specnum), , drop=FALSE]
  if (nrow(idswabs)>1) {
    d <- as.matrix(dist(idswabs$specdate))
    diag(d) <- Inf
    pairs <- which(d<100, arr.ind = TRUE)
    map2_dfr(pairs[,1], pairs[,2], function(i,j){
      if (idswabs$specdate[[i]]>idswabs$specdate[[j]]){
        temp <- i
        i <- j
        j <- temp
      }
      tibble(idA=idswabs$ID[[i]],
             idB=idswabs$ID[[j]],
             specA=idswabs$specnum[[i]],
             specB=idswabs$specnum[[j]],
             dateA=idswabs$specdate[[i]],
             dateB=idswabs$specdate[[j]],
             time_diff=idswabs$specdate[[j]]-idswabs$specdate[[i]])
    })
  } else {
    return(NULL)
  }
})

# remove duplicates
test_pairs <- test_pairs[!duplicated(test_pairs),]

#only keep pairs who's first case was at least 2 months after treatment
abx_summary$ID <- paste(abx_summary$codenum, "Infant", sep="-")
abx_summary$date_start <- lubridate::as_date(abx_summary$date_start)

test_pairs$prev_treatment <- map_lgl(1:nrow(test_pairs), function(i){
  sub_abx <- abx_summary[abx_summary$ID==test_pairs$idA[[i]], , drop=FALSE]
  keep <- sub_abx$date_start < test_pairs$dateA[[i]]
  sub_abx <- sub_abx[keep, , drop=FALSE]
  sub_abx <- sub_abx[!is.na(sub_abx$drug1),,drop=FALSE]
  if (nrow(sub_abx)<1){
    return(FALSE)
  }
  else if ((test_pairs$dateA[[i]]-max(sub_abx$date_start, na.rm = TRUE)) < 28){
    return(TRUE)
  } else {
    return(FALSE)
  }
})

#add treatment information
test_pairs$treatment <- map(1:nrow(test_pairs), function(i){
  sub_abx <- abx_summary[abx_summary$ID==test_pairs$idA[[i]], , drop=FALSE]
  keep <- (sub_abx$date_start >= test_pairs$dateA[[i]]) & (sub_abx$date_start < test_pairs$dateB[[i]])
  sub_abx <- sub_abx[keep, , drop=FALSE]
  sub_abx <- sub_abx[!is.na(sub_abx$drug1),,drop=FALSE]
  return(map_dfr(1:nrow(sub_abx), function(j){
    drugs <- unlist(sub_abx[j, c("drug1", "drug2", "drug3", "drug4")])
    drugs <- drugs[!is.na(drugs)]
    drugs <- map_dfr(drugs, ~ tibble(date=sub_abx$date_start[[j]], drug=.x))
  }))
})
test_pairs$is_treated <- map_lgl(test_pairs$treatment, ~ nrow(.x)>0)
test_pairs <- test_pairs[!test_pairs$prev_treatment,]

#remove those pairs where the follow up sample was too far from treatment
test_pairs$is_follow_up <- map_lgl(1:nrow(test_pairs), function(i){
  if (!test_pairs$is_treated[[i]]){
    return(TRUE)
  } else if ((test_pairs$dateB[[i]] - max(test_pairs$treatment[[i]]$date))<=28){
    return(TRUE)
  } else {
    return(FALSE)
  }
})
test_pairs <- test_pairs[test_pairs$is_follow_up,]

test_pairs <- test_pairs[order(test_pairs$specA, test_pairs$time_diff),]
test_pairs <- test_pairs[!duplicated(test_pairs$specA), ]
test_pairs <- test_pairs[order(test_pairs$specB, test_pairs$time_diff),]
test_pairs <- test_pairs[!duplicated(test_pairs$specB), ]

specs_to_keep <- (merged_serotype_calls %>% 
  filter(!is.na(consensus_percentage)) %>%
  filter(grepl("Infant", host)))$specnum


test_pairs <- test_pairs %>% 
  filter(specA %in% specs_to_keep)

test_pairs$pair <- paste(test_pairs$specA, test_pairs$specB, sep="_")
test_pairs$treatment <- map_chr(test_pairs$treatment, ~ paste(unlist(.x), collapse = "/"))
test_pairs$is_clearB <- swab_summary$pnc[match(test_pairs$specB, swab_summary$specnum)]=='No'
test_pairs$is_clearA <- swab_summary$pnc[match(test_pairs$specA, swab_summary$specnum)]=='No'
test_pairs$is_clearB <- test_pairs$is_clearB & !(test_pairs$specB %in% merged_serotype_calls$specnum)
test_pairs$is_clearA <- test_pairs$is_clearA & !(test_pairs$specA %in% merged_serotype_calls$specnum)

test_pairs <- test_pairs[grepl('.*AMOX.*', test_pairs$treatment) | (!test_pairs$is_treated),]
```

```{r}
transcluster <- fread('./data/combined_transcluster_probabilities.csv', header = TRUE) %>% as_tibble() %>%
  arrange(snp_distance) %>%
  filter(sampleA!=sampleB)
transcluster <- transcluster[!duplicated(map2_chr(transcluster$sampleA, transcluster$sampleB, ~{paste(sort(c(.x,.y)), collapse = '')})),]

transcluster$specnumA <- manifests_qc$`DONOR ID (required for EGA)`[match(gsub(".*-", "", transcluster$sampleA), manifests_qc$`Lane Name`)]
transcluster$specnumB <- manifests_qc$`DONOR ID (required for EGA)`[match(gsub(".*-", "", transcluster$sampleB), manifests_qc$`Lane Name`)]
transcluster$pair <- map2_chr(transcluster$specnumA, 
                             transcluster$specnumB, ~ 
                               paste(sort(c(.x,.y)), collapse = '-'))

transcluster <- transcluster %>% 
  group_by(pair) %>%
  summarise(
    min_snp = min(snp_distance),
    min_K0 = min(`K=0`)
  )

```


## Clearance probabilities

Estimate the maintenance of a PBP type after treatment. We require that we observe a type in at least 3 of both the treated and untreated pairs.

```{r}
CDC_resistance$PBP1A[CDC_resistance$PBP1A %in% c('NF','NEW')] <- NA
CDC_resistance$PBP2B[CDC_resistance$PBP2B %in% c('NF','NEW')] <- NA
CDC_resistance$PBP2X[CDC_resistance$PBP2X %in% c('NF','NEW')] <- NA

CDC_resistance$pbp_type <- paste(CDC_resistance$PBP1A, CDC_resistance$PBP2B, CDC_resistance$PBP2X)
pbp_types <- sort(table(CDC_resistance$pbp_type), decreasing = TRUE)
pbp_test_pairs <- test_pairs
pbp_test_pairs$typeA <- map(pbp_test_pairs$specA, function(a) CDC_resistance[CDC_resistance$specnum == a, 
                                                                             c('PBP1A', 'PBP2B', 'PBP2X')])
pbp_test_pairs$typeB <- map(pbp_test_pairs$specB, function(a) CDC_resistance[CDC_resistance$specnum == a, 
                                                                             c('PBP1A', 'PBP2B', 'PBP2X')])

pbp_test_pairs$GPSCA <- map(pbp_test_pairs$specA, ~ merged_serotype_calls$GPSC[merged_serotype_calls$specnum==.x])
pbp_test_pairs$GPSCB <- map(pbp_test_pairs$specB, ~ merged_serotype_calls$GPSC[merged_serotype_calls$specnum==.x])
pbp_test_pairs$cleared <- map2_lgl(pbp_test_pairs$GPSCA, pbp_test_pairs$GPSCB, ~{
  !any(.x %in% .y)
})

pbp_test_pairs$min_snp <- map2_dbl(pbp_test_pairs$specA, pbp_test_pairs$specB, ~{
  snp <-  transcluster$min_snp[transcluster$pair==paste(sort(c(.x,.y)), collapse = '-')]
  if (length(snp)<1) return(Inf)
  return(snp)
})

# define cleared as having no overlapping GPSCs and no plausible SNP distance <= 10
# we also consider cleared if the pbp type being considered is not found in the second sample of a pair
pbp_test_pairs$cleared <- pbp_test_pairs$cleared & (pbp_test_pairs$min_snp>10)

pbp_types <- unique(c(map(CDC_resistance$PBP1A, ~ c(1, .x)),
                      map(CDC_resistance$PBP2B, ~ c(2, .x)),
                      map(CDC_resistance$PBP2X, ~ c(3, .x))))
pbp_types <- pbp_types[map_lgl(pbp_types, ~ !is.na(.x[[2]]))]

tests_pbps <- map_dfr(pbp_types, function(type) {
  dat <- pbp_test_pairs
  dat$inA <- map_lgl(dat$typeA, ~{
    type[[2]] %in% unlist(.x[,as.numeric(type[[1]])])
  })
  dat$inB <- map_lgl(dat$typeB, ~{
    type[[2]] %in% unlist(.x[,as.numeric(type[[1]])])
  })
  
  dat$is_treated <- factor(dat$is_treated, levels = c(FALSE, TRUE))
  if (min(table(dat$is_treated[dat$inA]))<10) return(tibble())
  
  dat$cleared <- dat$cleared | ((!dat$inB) & dat$inA)

  m <- glm(cleared ~ time_diff + is_treated + inA + is_treated:inA, family = binomial, data = dat)
  conv <- m$converged
  m <- broom::tidy(m) %>% 
    filter(grepl('is_treatedTRUE:inATRUE', term)) %>%
    add_column(pbp_type=paste(c('PBP1A', 'PBP2B', 'PBP2X')[[as.numeric(type[[1]])]], type[[2]], collapse = ' - '), .before = 1)
  m$converged <- conv
  m$data <- list(dat)

  return(m)
}) %>% 
  filter(converged) %>%
  arrange(p.value)

tests_pbps$adj.p <- p.adjust(tests_pbps$p.value, method = 'BH')
top_pbps <- tests_pbps %>% filter(adj.p<0.1)
```


```{r}
tmp_out <- pbp_test_pairs[,c(1:7,10,12,17,18,19,20)]
tmp_out$GPSCA <- map_chr(tmp_out$GPSCA, ~ paste(.x, collapse = ','))
tmp_out$GPSCB <- map_chr(tmp_out$GPSCB, ~ paste(.x, collapse = ','))
write.table(tmp_out, 
            file = "~/Downloads/treatment_pairs.tsv",
            quote = FALSE, col.names = TRUE, row.names = FALSE, sep='\t')

tmp_out <- map_dfr(pbp_types, function(type) {
  dat <- pbp_test_pairs
  dat$inA <- map_lgl(dat$typeA, ~{
    type[[2]] %in% unlist(.x[,as.numeric(type[[1]])])
  })
  dat$inB <- map_lgl(dat$typeB, ~{
    type[[2]] %in% unlist(.x[,as.numeric(type[[1]])])
  })
  
  dat$is_treated <- factor(dat$is_treated, levels = c(FALSE, TRUE))
  if (min(table(dat$is_treated[dat$inA]))<10) return(tibble())
  
  dat$cleared <- dat$cleared | ((!dat$inB) & dat$inA)

  dat <- dat %>% add_column(PBP_type=paste(c('PBP1A', 'PBP2B', 'PBP2X')[as.numeric(type[[1]])], type[[2]], sep = '-'), .before = 1)

  return(dat)
})

tmp_out <- tmp_out[,-c(9,11,14,15,16,17)]

tmp_out$GPSCA <- map_chr(tmp_out$GPSCA, ~ paste(.x, collapse = ','))
tmp_out$GPSCB <- map_chr(tmp_out$GPSCB, ~ paste(.x, collapse = ','))
write.table(tmp_out, 
            file = "~/Downloads/pbp_treatment_pairs.tsv",
            quote = FALSE, col.names = TRUE, row.names = FALSE, sep='\t')
```

```{r}
gg1 <- ggplot(top_pbps, aes(x=pbp_type, y=estimate, col=estimate<0)) +
  geom_point(size=3) +
  geom_errorbar(aes(ymin=estimate-2*std.error, 
                    ymax=estimate+2*std.error), width=0.2, size=1) +
  geom_hline(yintercept=0, col='grey') +
  theme_clean(base_size = 20) +
    theme(plot.background = element_blank(),
          legend.background = element_blank(),
          strip.text.y = element_blank(),
          legend.position = 'none') +
  # theme(axis.text.x = element_text(angle = 45, hjust=1)) +
  scale_colour_manual(values = c('#0571b0', '#ca0020'), name='') +
  xlab('Type') +
  ylab('log odds ratio of clearance following treatment') +
  coord_flip()

gg1

ggsave("figures/pbpo_log_odds.png", width = 10, height = 7)
ggsave("figures/pbpo_log_odds.pdf", width = 10, height = 7)
```


```{r}
type_to_mic <- fread("~/Downloads/metcalf_Ref_PBPtype_MIC.csv") %>% as_tibble()
types <- str_split_fixed(type_to_mic$APT, '-', 3)
colnames(types) <- c('PBP1A', 'PBP2B', 'PBP2X')
type_to_mic <- cbind(type_to_mic,  types) %>% as_tibble()

pdf2 <- map_dfr(str_split(top_pbps$pbp_type, ' '), ~{
  tibble(type=paste(.x, collapse = ' '),
         mic=type_to_mic$AMO[.x[[2]]==unlist(type_to_mic[,.x[[1]]])])
})

gg2 <- ggplot(pdf2, aes(x=type, y=mic)) +
  geom_boxplot(outlier.colour = NA) +
  geom_point() +
  theme_clean(base_size = 20) +
    theme(plot.background = element_blank(),
          legend.background = element_blank(),
          strip.text.y = element_blank(),
          legend.position = 'none') +
  xlab('') + ylab('MIC') +
  coord_flip() 

gg1 + gg2


ggsave("figures/pbpo_log_odds_w_mic.png", width = 14, height = 7)
ggsave("figures/pbpo_log_odds_w_mic.pdf", width = 14, height = 7)
```

```{r}
single_colony_cdc <- fread("./data/single_colony_TABLE_SPN_CDC_pipeline_Typing_Results.txt", data.table = FALSE)
single_mash <- fread("./data/msweep_verification/combined_single_mash_results.tab", data.table = FALSE, 
                     col.names = c("sample", "identity", "shared-hashes", "median-multiplicity", "p-value", "query-ID","group")) %>% as_tibble()
single_lineage_poppunk <- fread("./data/poppunk_gpsc_definitive.csv") %>% as_tibble()
single_lineage_poppunk$sample <- gsub("GPS_all_assemblies/","",gsub("\\.contigs.*", "", single_lineage_poppunk$sample))


single_colony_cdc$GPSC <- single_lineage_poppunk$GPSC[match(single_colony_cdc$Sample, single_lineage_poppunk$sample)]

gpsc1_cdc <- single_colony_cdc %>% filter(GPSC==1)

table(gpsc1_cdc$PBP1A)
table(gpsc1_cdc$PBP2B)
table(gpsc1_cdc$PBP2X)
```

