---
title: "Initial Quality Control"
author: "Gerry Tonkin-Hill"
date: "`r Sys.Date()`"
output: 
  html_document:
    fig_width: 12
    fig_height: 8
editor_options: 
  chunk_output_type: console
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(fig.width=20, fig.height=12,
                      echo=TRUE, warning=FALSE, message=FALSE,
                      tidy=TRUE)
options(stringsAsFactors = FALSE)
```

This markdown file documents the initial filters to remove low quality samples.

##Load Libraries

```{r}
library(tidyverse)
library(ggplot2)
library(data.table)
library(ggthemes)
library(patchwork)
options(stringsAsFactors = FALSE)
```

Kraken was used for contamination detection.

Load data

```{r}
# load Kraken output
contam_df <- fread("./data/pipeline_qc/5000.pf_qc_contam_LS.csv")
contam_df <- as_tibble(data.table::melt(contam_df, 
                                        id.vars="Species", 
                                        variable.name="lane", 
                                        value.name="percentage", 
                                        variable.factor=FALSE))
```

```{r}
manifests <- fread("./data/manifests.csv", data.table = FALSE)
stats <- fread("./data/pipeline_qc/5000.pathfind_stats.csv", data.table = FALSE)
manifests <- merge(manifests, stats, by.x="SANGER SAMPLE ID", by.y = "Sample", all = TRUE)

# remove those not sequenced
manifests <- manifests[!is.na(manifests$`Lane Name`),]

# check the data frames are include everything
stopifnot(all(contam_df$lane %in% manifests$`Lane Name`))

# remove single isolate picks
manifests <- manifests[!grepl("-", manifests$`DONOR ID (required for EGA)`),]

# remove sequencing controls
manifests <- manifests[!grepl("control", manifests$`SAMPLE DESCRIPTION`),]
```

##Coverage

```{r}
gg1 <- ggplot(manifests, aes(x=`Depth of Coverage`)) + 
  geom_histogram(bins = 50) +
  theme_clean(base_size = 16) +
  theme(plot.background = element_blank(),
        legend.background = element_blank()) +
  xlab('depth of coverage') +
  geom_vline(xintercept = 50, col='red')
gg1

# initially we exclude those with less the 50 fold coverage
manifests <- manifests[manifests$`Depth of Coverage`>50,]
```

##Contamination

```{r}
contam_df <- contam_df[contam_df$lane %in% manifests$`Lane Name`,]

gg2 <- ggplot(contam_df %>% filter(Species=="Streptococcus pneumoniae"), aes(x=percentage)) + 
  geom_histogram(bins = 50) +
  theme_clean(base_size = 16) +
  theme(plot.background = element_blank(),
        legend.background = element_blank()) +
  xlab(expression(~italic('Streptococcus pneumonie %'))) +
  geom_vline(xintercept = 75, col='red') +
  ylab("")
gg2

gg1 + gg2 + patchwork::plot_layout(nrow = 1)

ggsave("./figures/coverage_contam_histogram.png", width = 12, height = 7)
ggsave("./figures/coverage_contam_histogram.pdf", width = 12, height = 7)

# we initially ignore those with less than 75% S. pneumo
contams <- unique(contam_df$lane[(contam_df$Species=="Streptococcus pneumoniae") & (contam_df$percentage<75)])
contam_df <- contam_df[contam_df$Species!="Total",]

contminants_df <- contam_df[(!contam_df$lane %in% contams) & (contam_df$percentage>1),]

species_order <- contminants_df %>%
  group_by(Species) %>%
  summarise(m=sum(percentage),
            count=n()) %>%
  arrange(m)
contminants_df$Species <- factor(contminants_df$Species, levels = species_order$Species)
contminants_df <- contminants_df[contminants_df$Species %in% species_order$Species[species_order$count>=5],]

ggplot(contminants_df, aes(y=Species, x=percentage)) + 
  geom_boxplot(outlier.color = NA) +
  geom_jitter(width = 0, height = 0.1, alpha=0.5) +
  theme_clean(base_size = 16) +
  theme(plot.background = element_blank(),
        legend.background = element_blank()) +
  xlab("percentage of sample") +
  ylab("species")

ggsave("./figures/contam_boxplots.png", width = 9, height = 7)
ggsave("./figures/contam_boxplots.pdf", width = 9, height = 7)
```

For now we take a moderate approach and ignore those with less than 75% *S. pneumoniae*. For fine scale variation we include additional filters in the post processing pipelines.

```{r}
contam_df <- contam_df[!(contam_df$lane %in% contams),]
manifests <- manifests[manifests$`Lane Name` %in% contam_df$lane,]
```

We now write out this moderate filtering to file for input to downstream analysis.

```{r}
penumo_percent <- contam_df[contam_df$Species=="Streptococcus pneumoniae",]
manifests$percent_pneumo <- penumo_percent$percentage[match(manifests$`Lane Name`, penumo_percent$lane)]
# write.csv(manifests, "./data/manifest_w_QC.csv", row.names = FALSE, quote = FALSE)
```













